GSS Education Analysis (Working Title)

### 2023-03-16 outputs/paper/paper.qmd
add introduction and update references

# Data and methods {#sec-data}

@@

This paper utilzes the R statistical programming language [@citeR], along with several packages, tidyverse [@citetidyverse], here [@citehere], janitor [@citejanitor],and dplyr [@citedplyr]. All figures in this paper have been created using the packages ggplot2 [@citeggplot2] and the tables have been created using knitr [@citeknitr] and kableExtra [@citekablextra]. The color styles in graphs have been created using the RColorBrewer packages [@citercolorbrewer].

@@

All data within this paper are extracted from the 2021 United States General Social Survey (GSS). This survey consists of a series of nationally representative cross-sectional interviews and collects data on contemporary American society in order to explain trends in opinions, attitudes and behaviors and monitor these trends **\[citegsscodebook\].** Since it began tracking trends in public opinion in 1972, the GSS has used in-person data collection as its primary method of data collection but in 2021, it moved to an address-based sampling method with a focus on web and web-based self-administered questionnaires **\[citegsscodebook\]**. The 2021 GSS was conducted from December 1, 2020 to May 3, 2021, and was processed in accordance with standard NORC procedures **\[citegsscodebook\].**

@@

The 2021 GSS samples adults over the age of 18 in the United States who do not live in institutional housing **\[citegsscodebook\].** This GSS was released in three batches, the first consisted of 10,091 addresses, the second consisted of 10,000 addresses and the third consisted of 7,500 addresses **\[citegsscodebook\]**. From these three releases, 1,271 completed the survey from the first batch, 1,391 completed the survey from the second batch, and 1,069 completed the survey from the third batch **\[citegsscodebook\].** Meaning that there was an overall response yield of 13.5% or 3,731 complete responses from the 27,591 total surveys released **\[citegsscodebook\].**

@@

This paper is assessing our estimand of 1,567 responses of the 3,731 complete responses received which represents 42% of the complete respondents and only 5.7% of the total surveys released. Respondent's to the 2021 GSS are males or female US citizens, not living in institutional housing, who are over the age of 18 **\[citegsscodebook\]**. In our analysis we removed 2,164 responses to account for unanswered or not applicable responses. We did this to ensure that the data used in this paper is complete and representative of all variables. The variables we selected for our analysis can be seen in @tbl-1 and their measurement levels can be seen in @tbl-2. Responses were measured using likert scales which measured respondents opinions to the questions being asked.

@@

## Bias and ethics {#sec-fifth-point}

It is important to note that observations from the 2021 GSS may be influenced by social trends due to the world around them, such as the COVID-19 pandemic along with methodological shifts that occurred for this survey. As the GSS changed to a full web-based survey in this year, answers could have been impacted. This is why we have chosen not to compare previous survey results. Additionally, as this survey was completed online, there is a higher rate of non-response. To control for this the 2021 GSS adjusted population totals to ensure that the weighted totals were able to match closely to the U.S Census Bureau estimates by sex, age, education, race, ethnicity, and  region of the country **\[citegsscodebook\]**. Additionally due to the online nature of the survey in 2021, household enumeration was not possible and so households had to identify the adult with the most birthday **\[citegsscodebook\]**. This could have resulted in respondent selections occurring at the household level or missing some household residents such as people living abroad, adult children, etc. **\[citegsscodebook\]**. More bias can occur from sampling error in which the sample conducted from this survey could not be representative of the US population **\[citegsscodebook\]**. Although, the US GSS does control for this by conducting tests of significance to look for trend differences estimates **\[citegsscodebook\].**

@@

The United States is built upon individualism, which trickles into their definitions of financial security, happiness, and overall fulfillment and satisfaction in life. Well-being of individuals within the United States therefore, is tracked and monitored through a variety of different factors. The aim of this investigation is to determine whether higher educational achievement influences respondent\'s perceptions of their social class, degree of financial satisfaction and job satisfaction. These factors were chosen because of their significance with regards to the American Dream, a concept that is built off a linear pattern of behavior in terms of economic security. Hustle culture is popular within the United States as an extension of the American Dream, because the idea that an individual must work hard in order to reap the benefits in terms is what drives culture within America.

According to Economic Security and the American Dream, the rewards of the American Dream included owning a home, having access to quality health care, having a job that enables an individual to support their family, and having a secure and stately retirement [@citerankhirschl]**.**These are the fundamentals in terms of how Americans generally define their well-being in society. However, no matter how linear an American\'s personal development or educational journey may be, it is simply impossible to be able to guarantee the rewards that the American Dream has to offer. But the ideals of living in the United States, has a powerful impact on culture, and how individuals view themselves and their position within society, but how they fare in comparison to those around them.

The assumption in general is a higher degree of education translates to an overall greater quality of life or well-being, greater job satisfaction, higher social class, etc. However, we found that this was not necessarily the case. Overall, out of our participants that completed the survey, the largest group had only completed high school, and the second largest group had completed a bachelor\'s degree. The level of education was compared to a few variables: self-ascribed social class, financial satisfaction,  job satisfaction, and financial ranking against other families in the US.

@@

However, in some cases, the positive correlation between higher education and overall subjective well-being is more prominent - resulting in significant tension in our data of consideration. Researchers have investigated the relationship between education and one\'s subjective well-being, with several studies indicating that a higher level of education may be positively associated with greater levels of happiness. This relationship may be explained by the fact that education can impact one\'s self-ascribed social class, which in turn can influence their notion of happiness. Higher education is assumed to provide individuals with better job opportunities, higher income, better access to resources, and greater social mobility - placing people on the path towards a successful American Dream and resulting in a higher sense of one\'s social status. As a result, these individuals might experience greater life satisfaction, which contributes to their overall sense of happiness and security.

In a 2020 article published in the Journal of Personality and Social Psychology, authors Siyu Yu and Steven L. Blader examined the impact of status and social class on subjective well-being (SWB). They found that often times social class addresses fundamental psychological needs, such as autonomy, mastery, respect, and occupational prestige, and they proposed that individuals in higher social classes experience greater levels of status and power, leading to higher SWB **[@citeyublader].** Conversely, individuals in lower social classes feel lower levels of status and power, resulting in lower levels of SWB. This insight is evident in @fig-3, where we found that individuals holding Bachelors and Graduate Degrees are proportionally more satisfied with their financial situation, with the large majority selecting \"Pretty Well Satisfied\" and \"More or Less Satisfied\" than those selecting \"Not Satisfied at All\". Out of the 426 responses indicating a \"Pretty Well\" financial satisfaction, 69% of responses were individuals holding Bachelor and Graduate degrees while only 31% held Associate/Junior college, highschool, and less than highschool levels of education. In comparison, those who only completed up until the high school level of education have the most responses of \"Not Satisfied at All\". 70% out of the 390 \"Not Satisfied At All\" responses were individuals who completed highschool education or lower, while only 30% of those holding Bachelor and Graduate degrees indicated a low financial satisfaction.

In addition, when looking at @fig-4, we found that there are more individuals holding Bachelor and Graduate degrees who have identified themselves as Upper and Middle class in proportion to those with lower levels of education. 89% of those who selected \"Upper Class\" and 68% of \"Middle Class\" held Bachelor and Graduate degrees. Among the 676 responses for \"Working Class\", 471 of those held high school level of education and lower, making up 70% of the responses. In sum, we found that both the research and data point to a link between higher social class and greater perceived levels of life satisfaction.

Overall, there seems to be a significant relationship between the data in education, social class, and one\'s perceived well-being. Education may serve as an important factor in determining an individual\'s social class as it can unlock opportunities for resources and mobility, which in turn can positively impact their sense of status and overall happiness. At the same time, while higher levels of education may be generally associated with greater levels of overall well-being, this relationship may not always be a cause and effect correlation. Subjective well-being may be further influenced by other factors such as personal values, cultural background, and social support. It is important to consider these important nuances when examining the data and making assumptions in order to develop a more comprehensive understanding of how variables interact and impact individuals\' lives. This will be further explored in the weaknesses, ethics, and bias section of this paper.

@@

The second largest group of participants who had completed the survey had completed a bachelor's degree. In general, while a higher level of education improves access to better job opportunities, the process is not as linear as it may seem. According to @fig-1, while the number of individuals who have a graduate degree is not as high as the group that has completed a bachelor\'s degree, the reported financial ranking against other families was proportionately very similar for both groups. The majority for both groups claimed to be average or above average. There was still a significant number of individuals who had chosen below average, even for the group with the highest level of education.

A similar notion exists when comparing an individuals level of education with their job satisfaction. According to @fig-2, as the education level rises so does the number of people who are proportionately more satisfied with their jobs. However, job satisfaction is fleeting, it is simply a reflection of the current state, and not an overall indication of cumulative job satisfaction. According to Figure 3, financial satisfaction rises with an increased level of education, but the group that claims dissatisfaction is a significant number in the bachelor\'s and graduate degree groups. It remains a minority, but it is significant nonetheless. This points towards the un-attainability of the American Dream, in that despite the greater level education and increased access that comes with it, it does not guarantee increased job satisfaction, or a higher financial satisfaction.

## Limitations and weaknesses {#sec-limitations-and-weaknesses}

There are several limitations to this paper. One weakness is that our insights rely on data from the General Social Survey (GSS), which is a cross-sectional survey that collects data at a single point in time. Although it can be helpful in investigating associations between variables and developing hypotheses, the GSS is not capable of establishing causation between variables and has potential for selection bias if participants are not representative of the population of interest [@citestrengthsweaknesses]. In addition, due to changes in the questions asked over time, the GSS has limited capacity to make exact comparisons - as in the data analyzed in our study, which was collected during a period when the survey was conducted entirely online, which necessitated modifications to the question format in order to accommodate the new platform. Above all, due to the nature of the data being self-reported responses, it may not fully capture the dynamic and complicated nature of the ways in which various levels of education affect one\'s overall well-being. The General Social Survey responses are subject to influence by a myriad of external factors including recall bias, in which participants may have left out important details when reporting about their subjective experiences.

Another potential ethical concern is that our paper may inadvertently reinforce the concept of the American Dream, which suggests that higher education is a key factor to achieving happiness and success in life. This concern is particularly relevant in @sec-social-class, where we analyze the data as-is without taking into account the presence of systemic inequalities such as racism and economic inequality that can significantly limit access to one\'s education and social mobility. Moreover, the sample survey in the GSS may not fully represent the experiences of more marginalized groups, which point to the need for further research and survey strategies that incorporate more diverse perspectives.

In sum, while our paper provides valuable insights on the relationship between education and one\'s perceived well-being, it is important to consider the limitations, ethical concerns, and potential biases in the data and interpretation of our insights and findings.

>>>>

In the United States, the concept of the American Dream has been deeply ingrained in the national consciousness for decades. It is the idea that through hard work and determination, anyone can achieve success and upward mobility regardless of their background and other significant, uncontrollable factors. Within this notion, education is often perceived as one of the key factors in achieving the American Dream, with many individuals in the United States believing that a degree is necessary for financial stability and social status. But does this belief reflect reality? We analyzed data from the 2021 United States General Social Survey to examine the relationships between education level, self-perceived social class, financial satisfaction, and job satisfaction and how they relate to an individual\'s overall sense of well-being.

We will first discuss the source of our data, its biases, and review the sampling and key features including our estimand of 1,567 responses of the 3,731 complete responses received from male and female US citizens, not living in institutional housing, and who are over the age of 18. Then, we will discuss our results through the lens of the American Dream, Social Class, and Financial Satisfaction.  Our findings were surprisingly complex and multifaceted, revealing that the relationships between these variables are not as straightforward as one might think.

A key insight that we found was that despite higher education (Bachelor and Graduate degrees) being positively associated with both financial satisfaction and self-perceived social class in the GSS data, the majority of respondents across all education levels still identified themselves as "working class" or "middle class\", suggesting an influence of the American idealism of \"hard work\". Moreover, our results indicated that job satisfaction levels varied significantly among respondents, regardless of their level of education -- indicating that higher education doesn't necessarily guarantee job satisfaction. Such findings raise significant tension between the ideal of the American Dream and its relationship to education. Our results suggest that the idea that a degree in higher education can put individuals towards a linear path to success and upward mobility is flawed.

Finally, we will take a look at the limitations and weaknesses of our findings in this paper and review the next steps on this particular topic. All in all, by examining the GSS survey data, we can shed light on the larger societal implications of the so-called American Dream, question existing assumptions and beliefs, and provide an insightful contribution to the ongoing discourse surrounding education and success in the United States.

# Data and Methods {#sec-data}

@@

This paper utilizes the R statistical programming language [@citeR], along with several packages, tidyverse [@citetidyverse], here [@citehere], janitor [@citejanitor],and dplyr [@citedplyr]. All figures in this paper have been created using the packages ggplot2 [@citeggplot2] and the tables have been created using knitr [@citeknitr] and kableExtra [@citekablextra]. The color styles in graphs have been created using the RColorBrewer packages [@citercolorbrewer].

@@

All data within this paper are extracted from the 2021 United States General Social Survey (GSS). This survey consists of a series of nationally representative cross-sectional interviews and collects data on contemporary American society in order to explain trends in opinions, attitudes and behaviors and monitor these trends [@citegsscodebook**].** Since it began tracking trends in public opinion in 1972, the GSS has used in-person data collection as its primary method of data collection but in 2021, it moved to an address-based sampling method with a focus on web and web-based self-administered questionnaires **[@citegsscodebook**]. The 2021 GSS was conducted from December 1, 2020 to May 3, 2021, and was processed in accordance with standard NORC procedures [@citegsscodebook]**.**

@@

The 2021 GSS samples adults over the age of 18 in the United States who do not live in institutional housing [@citegsscodebook]**.** This GSS was released in three batches, the first consisted of 10,091 addresses, the second consisted of 10,000 addresses and the third consisted of 7,500 addresses [@citegsscodebook]. From these three releases, 1,271 completed the survey from the first batch, 1,391 completed the survey from the second batch, and 1,069 completed the survey from the third batch [@citegsscodebook]**.** Meaning that there was an overall response yield of 13.5% or 3,731 complete responses from the 27,591 total surveys released [@citegsscodebook]**.**

@@

This paper is assessing our estimand of 1,567 responses of the 3,731 complete responses received which represents 42% of the complete respondents and only 5.7% of the total surveys released. Respondent's to the 2021 GSS are males or female US citizens, not living in institutional housing, who are over the age of 18 [@citegsscodebook]. In our analysis we removed 2,164 responses to account for unanswered or not applicable responses. We did this to ensure that the data used in this paper is complete and representative of all variables. The variables we selected for our analysis can be seen in @tbl-1 and their measurement levels can be seen in @tbl-2. Responses were measured using likert scales which measured respondents opinions to the questions being asked.

@@

## Bias and Ethics {#sec-fifth-point}

It is important to note that observations from the 2021 GSS may be influenced by social trends due to the world around them, such as the COVID-19 pandemic along with methodological shifts that occurred for this survey. As the GSS changed to a full web-based survey in this year, answers could have been impacted. This is why we have chosen not to compare previous survey results. Additionally, as this survey was completed online, there is a higher rate of non-response. To control for this the 2021 GSS adjusted population totals to ensure that the weighted totals were able to match closely to the U.S Census Bureau estimates by sex, age, education, race, ethnicity, and  region of the country [@citegsscodebook]. Additionally due to the online nature of the survey in 2021, household enumeration was not possible and so households had to identify the adult with the most birthday [@citegsscodebook]. This could have resulted in respondent selections occurring at the household level or missing some household residents such as people living abroad, adult children, etc. [@citegsscodebook]. More bias can occur from sampling error in which the sample conducted from this survey could not be representative of the US population [@citegsscodebook]. Although, the US GSS does control for this by conducting tests of significance to look for trend differences estimates [@citegsscodebook]**.**

@@

The United States is built upon individualism, which trickles into their definitions of financial security, happiness, and overall fulfillment and satisfaction in life. Well-being of individuals within the United States therefore, is tracked and monitored through a variety of different factors. The aim of this investigation is to determine whether higher educational achievement influences respondent's perceptions of their social class, degree of financial satisfaction and job satisfaction. These factors were chosen because of their significance with regards to the American Dream, a concept that is built off a linear pattern of behavior in terms of economic security. Hustle culture is popular within the United States as an extension of the American Dream, because the idea that an individual must work hard in order to reap the benefits in terms is what drives culture within America.

According to Economic Security and the American Dream, the rewards of the American Dream included owning a home, having access to quality health care, having a job that enables an individual to support their family, and having a secure and stately retirement [@citerankhirschl]**.**These are the fundamentals in terms of how Americans generally define their well-being in society. However, no matter how linear an American's personal development or educational journey may be, it is simply impossible to be able to guarantee the rewards that the American Dream has to offer. But the ideals of living in the United States, has a powerful impact on culture, and how individuals view themselves and their position within society, but how they fare in comparison to those around them.

The assumption in general is a higher degree of education translates to an overall greater quality of life or well-being, greater job satisfaction, higher social class, etc. However, we found that this was not necessarily the case. Overall, out of our participants that completed the survey, the largest group had only completed high school, and the second largest group had completed a bachelor's degree. The level of education was compared to a few variables: self-ascribed social class, financial satisfaction,  job satisfaction, and financial ranking against other families in the US.

@@

However, in some cases, the positive correlation between higher education and overall subjective well-being is more prominent - resulting in significant tension in our data of consideration. Researchers have investigated the relationship between education and one's subjective well-being, with several studies indicating that a higher level of education may be positively associated with greater levels of happiness. This relationship may be explained by the fact that education can impact one's self-ascribed social class, which in turn can influence their notion of happiness. Higher education is assumed to provide individuals with better job opportunities, higher income, better access to resources, and greater social mobility - placing people on the path towards a successful American Dream and resulting in a higher sense of one's social status. As a result, these individuals might experience greater life satisfaction, which contributes to their overall sense of happiness and security.

In a 2020 article published in the Journal of Personality and Social Psychology, authors Siyu Yu and Steven L. Blader examined the impact of status and social class on subjective well-being (SWB). They found that often times social class addresses fundamental psychological needs, such as autonomy, mastery, respect, and occupational prestige, and they proposed that individuals in higher social classes experience greater levels of status and power, leading to higher SWB [@citeyublader]**.** Conversely, individuals in lower social classes feel lower levels of status and power, resulting in lower levels of SWB. This insight is evident in @fig-3, where we found that individuals holding Bachelors and Graduate Degrees are proportionally more satisfied with their financial situation, with the large majority selecting "Pretty Well Satisfied" and "More or Less Satisfied" than those selecting "Not Satisfied at All". Out of the 426 responses indicating a "Pretty Well" financial satisfaction, 69% of responses were individuals holding Bachelor and Graduate degrees while only 31% held Associate/Junior college, high school, and less than high school levels of education. In comparison, those who only completed up until the high school level of education have the most responses of "Not Satisfied at All". 70% out of the 390 "Not Satisfied At All" responses were individuals who completed high school education or lower, while only 30% of those holding Bachelor and Graduate degrees indicated a low financial satisfaction.

In addition, when looking at @fig-4, we found that there are more individuals holding Bachelor and Graduate degrees who have identified themselves as Upper and Middle class in proportion to those with lower levels of education. 89% of those who selected "Upper Class" and 68% of "Middle Class" held Bachelor and Graduate degrees. Among the 676 responses for "Working Class", 471 of those held high school level of education and lower, making up 70% of the responses. In sum, we found that both the research and data point to a link between higher social class and greater perceived levels of life satisfaction.

Overall, there seems to be a significant relationship between the data in education, social class, and one's perceived well-being. Education may serve as an important factor in determining an individual's social class as it can unlock opportunities for resources and mobility, which in turn can positively impact their sense of status and overall happiness. At the same time, while higher levels of education may be generally associated with greater levels of overall well-being, this relationship may not always be a cause and effect correlation. Subjective well-being may be further influenced by other factors such as personal values, cultural background, and social support. It is important to consider these important nuances when examining the data and making assumptions in order to develop a more comprehensive understanding of how variables interact and impact individuals' lives. This will be further explored in the weaknesses, ethics, and bias section of this paper.

@@

The second largest group of participants who had completed the survey had completed a bachelor's degree. In general, while a higher level of education improves access to better job opportunities, the process is not as linear as it may seem. According to @fig-1, while the number of individuals who have a graduate degree is not as high as the group that has completed a bachelor's degree, the reported financial ranking against other families was proportionately very similar for both groups. The majority for both groups claimed to be average or above average. There was still a significant number of individuals who had chosen below average, even for the group with the highest level of education.

A similar notion exists when comparing an individuals level of education with their job satisfaction. According to @fig-2, as the education level rises so does the number of people who are proportionately more satisfied with their jobs. However, job satisfaction is fleeting, it is simply a reflection of the current state, and not an overall indication of cumulative job satisfaction. According to Figure 3, financial satisfaction rises with an increased level of education, but the group that claims dissatisfaction is a significant number in the bachelor's and graduate degree groups. It remains a minority, but it is significant nonetheless. This points towards the un-attainability of the American Dream, in that despite the greater level education and increased access that comes with it, it does not guarantee increased job satisfaction, or a higher financial satisfaction.

## Limitations and Weaknesses {#sec-limitations-and-weaknesses}

There are several limitations to this paper. One weakness is that our insights rely on data from the General Social Survey (GSS), which is a cross-sectional survey that collects data at a single point in time. Although it can be helpful in investigating associations between variables and developing hypotheses, the GSS is not capable of establishing causation between variables and has potential for selection bias if participants are not representative of the population of interest [@citestrengthsweaknesses]. In addition, due to changes in the questions asked over time, the GSS has limited capacity to make exact comparisons - as in the data analyzed in our study, which was collected during a period when the survey was conducted entirely online, which necessitated modifications to the question format in order to accommodate the new platform. Above all, due to the nature of the data being self-reported responses, it may not fully capture the dynamic and complicated nature of the ways in which various levels of education affect one's overall well-being. The General Social Survey responses are subject to influence by a myriad of external factors including recall bias, in which participants may have left out important details when reporting about their subjective experiences.

Another potential ethical concern is that our paper may inadvertently reinforce the concept of the American Dream, which suggests that higher education is a key factor to achieving happiness and success in life. This concern is particularly relevant in @sec-social-class, where we analyze the data as-is without taking into account the presence of systemic inequalities such as racism and economic inequality that can significantly limit access to one's education and social mobility. Moreover, the sample survey in the GSS may not fully represent the experiences of more marginalized groups, which point to the need for further research and survey strategies that incorporate more diverse perspectives.

In sum, while our paper provides valuable insights on the relationship between education and one's perceived well-being, it is important to consider the limitations, ethical concerns, and potential biases in the data and interpretation of our insights and findings.
